---
title: "RMS Titanic"
author: "branish"
date: 2020-7-16
output:
  github_document:
    toc: true
---

*Purpose*: Most datasets have at least a few variables. Part of our task in analyzing a dataset is to understand trends as they vary across these different variables. Unless we're careful and thorough, we can easily miss these patterns. In this challenge you'll analyze a dataset with a small number of categorical variables and try to find differences among the groups.

*Reading*: (Optional) [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)

df_titanic <- as_tibble(Titanic)
```

*Background*: The RMS Titanic sank on its maiden voyage in 1912; about 67% of its passengers died.

# First Look
<!-- -------------------------------------------------- -->

__q1__ Perform a glimpse of `df_titanic`. What variables are in this dataset?

```{r q1-task}
## TASK: Perform a `glimpse` of df_titanic
glimpse(df_titanic)
```

**Observations**:

- Variables: Class, Sex, Age, Survived, n

__q2__ Skim the [Wikipedia article](https://en.wikipedia.org/wiki/RMS_Titanic) on the RMS Titanic, and look for a total count of passengers. Compare against the total computed below. Are there any differences? Are those differences large or small? What might account for those differences?

```{r q2-summary}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_titanic %>% summarize(total = sum(n))
```

**Observations**:

- The Wikipedia article says there were approximately 1317 passengers and 885 crew members, for a total of around 2202, and that the ship was under capacity on its maiden voyage. Wikipedia's passenger count has one extra passenger compared to the dataset.
- Wikipedia says that the exact number of people aboard is not known because some ticketholders never boarded the ship, and some left before the end of the voyage.

```{r}
df_titanic %>% group_by(Class) %>% summarize(total_dataset = sum(n))
```
- The dataset has 325 1st-class passengers. Wikipedia has 324 1st-class passengers.
- 2nd class: the dataset has 285, Wikipedia has 284.
- 3rd class: the dataset has 706, Wikipedia has 709.
- This dataset does have 885 crew. This matches Wikipedia.

```{r}
df_titanic %>% group_by(IsCrewMember = (Class == "Crew"), Sex) %>% summarize(total_dataset = sum(n))
```
- The number of male and female passengers matches Wikipedia (869 male and 447 female). Wikipedia's extra passenger does not seem to be included in these totals. The Wikipedia article must have used different sources for the sex breakdown vs. the class breakdown.
- The number of male and female crew members matches Wikipedia (862 male and 23 female crew).

```{r}
df_titanic %>% group_by(Age) %>% summarize(total = sum(n))
```
- Dataset has 109 children (all passengers). Wikipedia has 107 children.

__q3__ Create a plot showing the count of passengers who *did* survive, along with aesthetics for `Class` and `Sex`. Document your observations below.

*Note*: There are many ways to do this.

```{r q3-task}
## TASK: Visualize counts against `Class` and `Sex`
df_titanic %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge") +
  facet_wrap( ~ Age)

df_titanic %>%
  filter(Survived == "Yes") %>%
  count(Sex, Class) %>%
  ggplot() +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge")

```

**Observations**:

- In first and second class, there were more female than male survivors.
```{r}
df_titanic %>%
  filter(Survived == "Yes", Class %in% c("1st", "2nd")) %>%
  count(Sex, Class) %>%
  ggplot() +
  ggtitle("Survivors in 1st and 2nd class") +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge")
```

- In 3rd class the numbers of male and female survivors were nearly equal.
```{r}
df_titanic %>%
  filter(Survived == "Yes", Class == "3rd") %>%
  count(Sex) %>%
  ggplot() +
  ggtitle("Survivors in 3rd class") +
  geom_col(mapping = aes(x = "Survivors in 3rd class", y = n, fill = Sex), position = "dodge")
```

- Among the crew, there were many more male survivors than female. However, there were only 23 female crew members in the first place.
- Sex ratios among the surviving children were much closer to even.
```{r}
df_titanic %>%
  filter(Survived == "Yes", Age == "Child") %>%
  ggplot() +
  ggtitle("Surviving children") +
  geom_col(mapping = aes(x = Class, y = n, fill = Sex), position = "dodge")
```

- There were not many surviving children in any passenger class, and none in crew. (There were no child crew members.)

# Deeper Look
<!-- -------------------------------------------------- -->

Raw counts give us a sense of totals, but they are not as useful for understanding differences between groups. This is because the differences we see in counts could be due to either the relative size of the group OR differences in outcomes for those groups. To make comparisons between groups, we should also consider *proportions*.[1]

The following code computes proportions within each `Class, Sex, Age` group.

```{r compute-prop}
## NOTE: No need to edit! We'll cover how to
## do this calculation in a later exercise.
df_prop <-
  df_titanic %>%
  group_by(Class, Sex, Age) %>%
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()
df_prop

df_prop_ageless <-
  df_titanic %>%
  group_by(Class, Sex, Survived) %>%
  summarize(n = sum(n)) %>%
  # now it's grouped by Class, Sex
  mutate(
    Total = sum(n),
    Prop = n / Total
  ) %>%
  ungroup()

  

```

__q4__ Replicate your visual from q3, but display `Prop` in place of `n`. Document your observations, and note any new/different observations you make in comparison with q3.

```{r q4-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  ggtitle("Proportion of passengers who survived by age, class, sex") +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
  facet_wrap( ~ Age)

df_prop_ageless %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  ggtitle("Proportion of passengers who survived by age, class") +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge")
```

**Observations**:

```{r}
df_prop %>%
  filter(Survived == "Yes", Age == "Child") %>%
  ggplot() +
  ggtitle("Proportion of children who survived by class, sex") +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge")
```
- All of the children in 1st and 2nd class survived.
- Less than half of the children in 3rd class survived. The survival rate for girls in 3rd class was higher than for boys in 3rd class.

```{r}
df_prop %>%
  filter(Survived == "Yes", Age == "Adult") %>%
  ggplot() +
  ggtitle("Proportion of adults who survived by class, sex") +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge")
```

- The survival rate for adult female passengers in 1st class was almost 100%. For adult males it was around a third of that.
- The survival rate for adult female passengers was almost 80% in 2nd class. For adult males it was closer to 10%.
- 80% of female crew members survived. Around 20% of male crew members survived.
- The survival rate for adult female passengers in 3rd class was under 50%. For adult males it was closer to 15%.


__q5__ Create a plot showing the group-proportion of passengers who *did* survive, along with aesthetics for `Class`, `Sex`, *and* `Age`. Document your observations below.

*Hint*: Don't forget that you can use `facet_grid` to help consider additional variables!

```{r q5-task}
df_prop %>%
  filter(Survived == "Yes") %>%
  ggplot() +
  ggtitle("Proportion of passengers who survived by age, class, sex") +
  geom_col(mapping = aes(x = Class, y = Prop, fill = Sex), position = "dodge") +
  facet_wrap( ~ Age)

```

**Observations**:

(See above)

# Notes
<!-- -------------------------------------------------- -->

[1] This is basically the same idea as [Dimensional Analysis](https://en.wikipedia.org/wiki/Dimensional_analysis); computing proportions is akin to non-dimensionalizing a quantity.
