---
title: "Gapminder"
author: "(Your name here)"
date: 2020-
output:
  github_document:
    toc: true
---

*Purpose*: Learning to do EDA well takes practice! In this challenge you'll further practice EDA by first completing a guided exploration, then by conducting your own investigation. This challenge will also give you a chance to use the wide variety of visual tools we've been learning.

<!-- include-rubric -->
# Grading Rubric
<!-- -------------------------------------------------- -->

Unlike exercises, **challenges will be graded**. The following rubrics define how you will be graded, both on an individual and team basis.

## Individual
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Effort | Some task __q__'s left unattempted | All task __q__'s attempted |
| Observed | Did not document observations | Documented observations based on analysis |
| Supported | Some observations not supported by analysis | All observations supported by analysis (table, graph, etc.) |
| Code Styled | Violations of the [style guide](https://style.tidyverse.org/) hinder readability | Code sufficiently close to the [style guide](https://style.tidyverse.org/) |

## Team
<!-- ------------------------- -->

| Category | Unsatisfactory | Satisfactory |
|----------|----------------|--------------|
| Documented | No team contributions to Wiki | Team contributed to Wiki |
| Referenced | No team references in Wiki | At least one reference in Wiki to member report(s) |
| Relevant | References unrelated to assertion, or difficult to find related analysis based on reference text | Reference text clearly points to relevant analysis |

## Due Date
<!-- ------------------------- -->

All the deliverables stated in the rubrics above are due on the day of the class discussion of that exercise. See the [Syllabus](https://docs.google.com/document/d/1jJTh2DH8nVJd2eyMMoyNGroReo0BKcJrz1eONi3rPSc/edit?usp=sharing) for more information.

```{r setup}
library(tidyverse)
library(gapminder)
library(knitr)
```

*Background*: [Gapminder](https://www.gapminder.org/about-gapminder/) is an independent organization that seeks to education people about the state of the world. They promote a "fact-based worldview" by focusing on data. The dataset we'll study in this challenge is from Gapminder.

# Guided EDA
<!-- -------------------------------------------------- -->

First, we'll go through a round of *guided EDA*. Try to pay attention to the high-level process we're going through---after this guided round you'll be responsible for doing another cycle of EDA on your own!

__q0__ Perform your "first checks" on the dataset. What variables are in this
dataset?

```{r q0-task}
## TASK: Do your "first checks" here!
glimpse(gapminder)
```

**Observations**:

- Write the variable names here

__q1__ Determine the most and least recent years in the `gapminder` dataset.

```{r q1-task}
## TASK: Find the largest and smallest values of `year` in `gapminder`
year_max <- summarize(gapminder, max(year))[1]
year_min <- summarize(gapminder, min(year))[1]
```

Use the following test to check your work.

```{r q1-tests}
## NOTE: No need to change this
assertthat::assert_that(year_max %% 7 == 5)
assertthat::assert_that(year_max %% 3 == 0)
assertthat::assert_that(year_min %% 7 == 6)
assertthat::assert_that(year_min %% 3 == 2)
print("Nice!")
```

__q2__ Filter on years matching `year_min`, and make a plot of the GDE per capita against continent. Choose an appropriate `geom_` to visualize the data. What observations can you make?

You may encounter difficulties in visualizing these data; if so document your challenges and attempt to produce the most informative visual you can.

```{r q2-task}
## TASK: Create a visual of gdpPercap vs continent
early <-
  gapminder %>%
  filter(year %in% year_min) %>%
  mutate(gdp = gdpPercap * pop)
```

```{r q2-continents}
early %>%
  select(continent, pop, gdp) %>%
  group_by(continent) %>%
  summarize(gdpTotal = sum(gdp), popTotal = sum(pop)) %>%
  mutate(gdpPercap = gdpTotal / popTotal) %>%
  ggplot() +
  ggtitle("GDP per capita for each continent - 1952") +
  geom_col(aes(x = continent, y = gdpPercap))
```
I interpreted the prompt as a request to plot the gdp per capita for each continent. The challenge then was to calculate the GDP per capita by dividing the sum of all GDPs on the continent by the total population of the continent.

However, there can't be three outliers among only five continents. I'll go back to looking at individual countries to find some outliers.


```{r q2-gdppercap}
early %>%
  ggplot() +
  ggtitle("Distribution of GDP per capita for each country in a continent - 1952") +
  geom_boxplot(aes(x = continent, y = gdpPercap))
```
The outlier in Asia is compressing all the other data points into the bottom of the plot. What country is that?
```{r q2-gdppercap-top}
top3GdpPercap <- 
  early %>%
  select(country, gdpPercap) %>%
  arrange(desc(gdpPercap)) %>%
  slice(1:3) %>%
  mutate(rank = rank(desc(gdpPercap)))
top3GdpPercap <- top3GdpPercap[ , c('rank', 'country', 'gdpPercap')]
topGdpPercap <- slice(top3GdpPercap, 1)
kable(top3GdpPercap, caption = 'Top 3 countries ranked by GDP per capita in 1952')
```
It's Kuwait.

```{r q2-gdppercap-2}
early %>%
  filter(!(country %in% topGdpPercap$country)) %>%
  ggplot() +
  ggtitle("Distribution of GDP per capita for each country in a continent - 1952") +
  labs(subtitle = "excluding Kuwait") +
  geom_boxplot(aes(x = continent, y = gdpPercap))
```

Now I'll look at GDP and population.
```{r q2-outliers}
# Top 3 countries ranked by GDP
top3gdp <- 
  early %>%
  select(country, gdp) %>%
  arrange(desc(gdp)) %>%
  slice(1:3) %>%
  mutate(rank = rank(desc(gdp)))
top3gdp <- top3gdp[ , c('rank', 'country', 'gdp')] # How do I reorder columns in the same pipeline/expression as the mutate?
topGdp = top3gdp %>% slice(1)
kable(top3gdp, caption = 'Top 3 countries ranked by GDP in 1952')

# Top 3 countries ranked by population
top3pop <- 
  early %>%
  select(country, pop) %>%
  arrange(desc(pop)) %>%
  slice(1:3) %>%
  mutate(rank = rank(desc(pop)))
top3pop <- top3pop[ , c('rank', 'country', 'pop')]
topPop = top3pop %>% slice(1:2)
kable(top3pop, caption = 'Top 3 countries ranked by population in 1952')
```


```{r q2-gdp}
early %>%
  ggplot() +
  ggtitle("Distribution of GDP for each country in a continent - 1952") +
  geom_boxplot(aes(x = continent, y = gdp))

 early %>%
   filter(!(country %in% topGdp$country)) %>%
   ggplot() +
   ggtitle("Distribution of GDP for each country in a continent - 1952") +
   labs(subtitle = "excluding United States") +
   geom_boxplot(aes(x = continent, y = gdp))
```


```{r q2-pop}
early %>%
  ggplot() +
  ggtitle("Distribution of country populations in each continent - 1952") +
  geom_boxplot(aes(x = continent, y = pop))

early %>%
  filter(!(country %in% top3pop$country)) %>%
  ggplot() +
  ggtitle("Distribution of country populations in each continent - 1952") +
  labs(subtitle = "excluding China and India") +
  geom_boxplot(aes(x = continent, y = pop))

early %>%
  filter(!(country %in% topGdpPercap$country), !(country %in% topPop$country)) %>%
  ggplot() +
  ggtitle('GDP per capita vs. population for countries in each continent - 1952') +
  labs(subtitle = 'excluding Kuwait, China, and India') +
  geom_point(aes(x = pop, y = gdpPercap)) +
  facet_wrap( ~ continent)


early %>%
  #filter(!(country %in% topGdpPercap$country)) %>%
  ggplot() +
  ggtitle('GDP per capita vs. population for all countries in dataset - 1952') +
  #labs(subtitle = 'excluding Kuwait') +
  geom_point(aes(x = pop, y = gdpPercap, shape = continent), size = 2)

```

**Observations**:

- Write your observations here

**Difficulties & Approaches**:

- If you try to plot gdpPerCap vs continent, geom_col will automatically try to add up the gdpPerCap of every country without taking into account the populations of each country. You need to calculate the GDP for each country, the GDP and population of each continent, and the GDP per capita of each continent manually.

__q3__ You should have found at least three outliers in q2. Identify those outliers (figure out which countries they are).

```{r q3-task}
## TASK: Identify the outliers from q2
kable(topPop, caption = "Highest population")
kable(topGdpPercap, caption = "Highest GDP per capita")
```

**Observations**:

- Kuwait (GDP per capita), China and India (population)

__q4__ Create a plot similar to yours from q2 studying both `year_min` and `year_max`. Find a way to highlight the outliers from q3 on your plot. Compare the patterns between `year_min` and `year_max`.

*Hint*: We've learned a lot of different ways to show multiple variables; think about using different aesthetics or facets.

```{r q4-task}
## TASK: Create a visual of gdpPercap vs continent
late <-
  gapminder %>%
  filter(year %in% year_max)

outliers <-
  topGdpPercap %>%
  full_join(topPop) %>%
  select(country)

both <- full_join(early, late)
bothoutliers <- both %>% filter(country %in% outliers$country)

both %>%
  ggplot() +
  ggtitle('GDP per capita vs. population for all countries in dataset') +
  geom_point(aes(x = pop, y = gdpPercap, shape = continent), size = 2) +
  geom_point(data = bothoutliers, aes(x = pop, y = gdpPercap, shape = continent, color = country), size = 2) +
  facet_wrap( ~ year)

both %>%
  filter(!(country %in% outliers$country)) %>%
  ggplot() +
  ggtitle('GDP per capita vs population for all countries EXCEPT Kuwait, China, India') +
  geom_point(aes(x = pop, y = gdpPercap, shape = continent), size = 2) +
  facet_wrap( ~ year)

bothoutliers %>%
  filter(country == "Kuwait") %>%
  ggplot() +
  ggtitle("GDP per capita vs population for Kuwait in 1952 and 2007") +
  geom_line(aes(x = pop, y = gdpPercap, group = country), size = .5, linetype = "longdash") +
  geom_point(aes(x = pop, y = gdpPercap, shape = factor(year)), size = 3) +
  expand_limits(x = 0, y = 0)

bothoutliers %>%
  filter(country %in% c("China", "India")) %>%
  ggplot() +
  ggtitle("GDP per capita vs population for China and India in 1952 and 2007") +
  geom_line(aes(x = pop, y = gdpPercap, color = country, group = country), size = .5, linetype = "longdash") +
  geom_point(aes(x = pop, y = gdpPercap, color = country, shape = factor(year)), size = 3) +
  expand_limits(x = 0, y = 0)


```

**Observations**:

* Kuwait's population grew by an order of magnitude, yet its GDP per capita only fell by a little more than 50%.
* China and India saw a severalfold increase in both GDP and population. In particular, China's population (more than) doubled, while its GDP per capita rose by an order of magnitude.
* I wonder if these GDP figures are adjusted for inflation.

# Your Own EDA
<!-- -------------------------------------------------- -->

Now it's your turn! We just went through guided EDA considering the GDP per capita at two time points. You can continue looking at outliers, consider different years, repeat the exercise with `lifeExp`, consider the relationship between variables, or something else entirely.

__q5__ Create *at least* three new figures below. With each figure, try to pose new questions about the data.

```{r q5-task1}
## TASK: Your first graph
usacanada <- gapminder %>% filter(country %in% c("United States", "Canada"))
gapminder %>%
  filter(continent == "Americas", !(country %in% c("United States", "Canada"))) %>%
  ggplot() +
  geom_line(aes(x = year, y = gdpPercap, color = country)) +
  #geom_line(data = usacanada, aes(x = year, y = gdpPercap, linetype = country), size = 2) +
  geom_line(data = filter(gapminder, country == "Argentina"), aes(x = year, y = gdpPercap, color = country), size = 2)
```
Q: Four countries in Central + South America saw a notable dip in GDP per capita in 2002, but then recovered by 2007. What happened and why did it only affect those countries?

```{r q5-task2}
## TASK: Your second graph
```
Sorry this isn't finished, I've fallen a couple days behind.

```{r q5-task3}
## TASK: Your third graph
```

